\vspace{-2.0cm}Stuff

### 2014-06-04 report/draft3.tex
Final Draft of Master's Project Report.

With Line Transects, a (straight) line is randomly placed in the study area, then walked by the observer, with any objects of interest noted and the distance perpendicular to the transect recorded. With point transects, the observer stands in one spot, recording the direct line distance from herself to the object of interest as if it was projected on the ground, within the 360\textdegree. This results in data similar to the line transect, but with more safety for the observer as they do not have to watch for moving wildlife at the same time as navigating potentially treacherous terrain. \cite{ramsey1979}.

With line transects, animals may react to an observer moving through the area by hiding or moving away. If the observer is standing still, and has allowed a ``cooling'' period for things to return to normal--possible with point transects--animal behavior may be more natural and less influenced by the presence of the observer.

@@

An estimate of population density is $n$, the count of objects observed, divided by the area in which they were observed, times the probability of observing the object at that distance. \cite{buckland2001} %eq 2.12, pg 39

$\hat{D}=n/(\mbox{Area})*P(\mbox{observing object}|\mbox{distance }r)$

@@

Figure~\ref{fig:82dist} illustrates the distance data for the Collared Kingfisher. This data represents the aggregated data from two islands, Tinian and Rota, and all four observers. We can observe several things in this plot:

@@

Secondly, the further away from the station, the distance observations were heaped into convenient distances. With smaller bin widths, this is also evident in the 100--200 m range as well.

Lastly, note that the detection distances are high between 100--200 m, and recall that the stations were placed 150 m apart. Since the stations on a transect were generally visited in sequence, this means it is possible a bird could have been observed from more than one station. How does this affect the independence of our observations? Will it make a difference in our final population density estimates?

\subsection{Independence}
Both \textcite{ramsey1979,buckland1987} discuss the assumption that the VCPs are placed randomly through the study area. This would seem to implicitly state that their observational ranges might by chance overlap, meaning a single object of interest could be observed from more than one station.  \textcite[240]{thompson2012} likewise indicates random placement of line transects which would seem to allow for a potential overlap of observation zones.

@@

\textcite[233]{buckland2001} discuss that each line or point must be ``randomly and independently located,'' with ``an equal probability of selection'' for all portions of the study area.

\textcite[235]{buckland2001} states ``Transects are normally spaced at a sufficient distance to avoid detecting an object from two neighboring transacts, although this is not usually critical unless sampling a line changes the animal distribution at neighboring, as yet unsampled lines.''

@@

In discussing line transects, \textcite{barry2001} treats it as a two-stage sample, with non-unique clusters (each study object can be in more than one cluster) having an equal probability of selection. This would seem to (but does not explicitly) endorse the possibility of overlapping VCPs.

One of the assumptions of distance sampling is that the objects being studied are distributed homogeneously with some uniform density, $D$, throughout the space \textcite{ramsey1981}.  (If the data is clustered we have other problems, not addressed here.) If the uniform distribution is the case, then it would make sense that it should not matter if the observations overlap, as long as the act of observing at one station does not affect the observations at the neighboring station. This echoes Buckland’s statement a few paragraphs previous.

\textcite[244]{thompson2012} mentions that the systematic selection of transects (random position of first, with additional transects a fixed distance apart) does not affect the approximate unbiasedness of the density estimator, but will have an effect on the approximate unbiasedness of the variance estimator.

It is clear the literature is divided on whether or not our observation areas should overlap. VCP layouts similar to \textcite{micronesian} ``random-systematic'' layout are common in the literature.

\subsection{Bias \& Consistency}
\textcite{buckland2001,ramsey1979} Buckland et al. (2001) and Ramsey \& Scott (1970) both claim that if our detectability estimate $g(z)=$1 out to $r$, then $\hat{D}$ is unbiased.

@@

	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/layout_rand-sys-4.pdf}
		\caption{A Transect Layout}
		\label{fig:transect1}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/layout_rand-sys-6.pdf}
		\caption{Another Transect Layout}
		\label{fig:transect2}
	\end{subfigure}
	\caption{VCP Layout options. On graph, 1 unit = 1 km. Circles represent 200 m mark.}
	\label{fig:layouts}
\end{figure}
Figures~\ref{fig:structured} and \ref{fig:random} illustrate the first two arrangements. Density in these images is based on the 20 birds per km$^2$ of the Palie region on Rota.

As seen in Figure~\ref{fig:82dist}, there is very little detection happening beyond 500 m. There are a total of 15 observations beyond 500 m, which is 0.9\% of the data.

@@

@@

Using the \texttt{hist()} function in R, the 1982 observation distances were broken into counts by 5 m groups. Using the bin midpoints, the density estimate was multiplied by 5 to get the density value for that bin. A LOESS regression was run with the midpoint as $x$ and $y=density*5$, using the \texttt{loess()} package in R with a span of 0.2. This is the solid line in Figure~\ref{fig:by5}.

This gives us an object that can be used with \texttt{predict()} to get density estimates for specific distances. The values generated by feeding our midpoints into our prediction function are represented by the dashed line in Figure~\ref{fig:by5}.

Following the scaling of the highest point to 1, as was done with the half normal curve, where the highest point was $g(0)$, we also find $\delta$ for this empirical detection probability. A sequence of $x$ values was generated covering the range from 0--500 m. This was plugged into the \texttt{predict()} function using our LOESS object. The highest density value was then scaled to be 1:

@@

\subsubsection{Movement}
In the Figure~\ref{fig:82dist}, the histogram of empirical observation distances, there appears to be evidence of movement away from the observer. To see if the movement would have an effect on the independence of our VCPs, depending on placement, a second simulation was run.

Three movement classes were defined:
\begin{enumerate}
\item No Movement: Objects stayed at original point.
\item Temporary Movement: Objects within 100 m had a random chance to move directly away from observer. Movement was larger if they started closer to observer. They were reset to original location before moving to next VCP
\item Permanent Movement: As with temporary movement, only not reset before moving on to next VCP. (Movement could compound if object was within the maximum observation radius for more than one VCP.)
\end{enumerate}

@@

\section{Results}
Two simulations were run, one looking at the three layout options crossed with the two detection probability function options and the other examining the effect of the three layout options against the three movement options, holding the detection probability option standard.
\subsection{Half-normal vs. Empirical $g(r)$}
\begin{figure}

@@

\end{figure}
Figure \ref{fig:sim1} shows the results of 1000 simulations, each analyzed with Empirical and Half-normal detection function, $g(x)$, and using each of the three layouts, Structured, Random, and Transect. All $\hat{D}$  were done using the kernel method with a normal kernel \parencite{quang1993}.

@@

As we would expect, the density estimates using $g(x)$ based on the Empirical distance observations is biased low. We also see bias in the estimates using the half-normal function for $g(x)$, but it is much less severe than in the Empirical data. \textcite{quang1993} does provide a bias-correction method, but it was not used for this simulation.

@@

Table~\ref{table:sim1} supports what we observe in the histogram: the centers vary between our two $g(r)$ options, but are approximately the same within the each option, regardless of which VCP layout we have chosen. Table~\ref{table:sim1capture} shows the percentage of simulations that captured the true object density of 20 per km$^2$ within a 95\% confidence interval. \cite{quang1993}

There may be evidence of a drop in accuracy with the Transect layout, but it is difficult to determine if this is a valid observation, or an artifact of this particular simulation. The drop, if it exists, is much less severe with the half-normal detection function. With the empirical detection function, the structured layout percentage swings much higher than the result from our random layout, unlike for half-normal where they are just over 1\% apart. Repeated simulations, could help narrow down if this is a true effect, or simply random noise.
\begin{table}[h]

@@

\end{figure}
\begin{table}[h]

@@

For testing combinations of movement and VCP layout, only the half-normal detection function was utilized. Figure~\ref{fig:sim} compares the histograms of the nine combinations. On the whole, the estimates are biased low, as was observed in the previous simulation, without movement. Within a column, the results are fairly similar in shape and spread. The most obvious differences lie between columns, indicating that the type of movement causes more bias in our estimate than the potential overlap of our observation areas.

The combination of compounded movement and the transect layout seems to have an effect on the density estimate. The other estimates for either movement type, are all in the same ballpark, around 11.8 with an approximate 59\% rate for capturing the true population value. For the Transect Layout and Compounded movement, this rises 0.50 to a mean population density estimate of 12.42, and a capture percent of 65\%. It is unclear if there is something about this combination that is reducing bias (possible, but feels unlikely), or if it is instead biasing the estimate upwards, which appears like bias reduction because the estimate is already low.

@@

\section{The Problem with Circles}
To test the effectiveness of the movement algorithm, the detection distances from the data where the objects moved was compared to the detection distances from the same data, assuming the objects stayed still. The movement data, (left panel, Figure~\ref{fig:moveTest}) is squashed more towards 0.1 Km than the still data (right panel). But there is a problem. The still data does not show the nice shoulder at 0 that the literature instructs us to expect.

@@

The take-away point is that for a VCP survey, even if everything is perfect, the empirical detection distances will not show the nice shouldered layout of the ideal detection probability curve. \textcite{ramsey1981} speak about this, when comparing the ``CUM-D'' curve, or the plot of the cumulative number of observations at distance $r$ for both line and point transects. However, almost all the illustrations on how the collected data should look, shows data that echoes the detection probability curve: the half-bell-shape with a nice shoulder. Since this shape is impossible for VCP, explicit examples should be shown in the literature so researchers do not assume their is something wrong with their data.

@@

There is a possible effect of the combination of compounded movement and the transect layout. It is unclear from these results what the particular cause of the increased estimate and accuracy is. Future work might involve modification of the simulation functions to return more information, such as the count of observed animals and the observation distances to assist with understanding why this combination is producing an estimate that appears out of line with it's neighbors.

@@

\printbibliography
%\bibliographystyle{plain}
%\bibliography{VCPbibliography}
\end{document}

>>>>

With Line Transects, a (straight) line is randomly placed in the study area, then walked by the observer, with any objects of interest noted and the distance perpendicular to the transect recorded. With point transects, the observer stands in one spot, recording the direct line distance from herself to the object of interest as if it was projected on the ground, within the 360\textdegree. This results in data similar to the line transect, but with more safety for the observer as they do not have to watch for moving wildlife at the same time as navigating potentially treacherous terrain. \parencite{ramsey1979}.

With line transects, animals may react to an observer moving through the area by hiding or moving away. If the observer is standing still, and has allowed a ``cooling'' period for things to return to normal---possible with point transects---animal behavior may be more natural and less influenced by the presence of the observer.

@@

An estimate of population density is $n$, the count of objects observed, divided by the area in which they were observed times the probability of observing the object at that distance. \cite{buckland2001} %eq 2.12, pg 39

$\hat{D}=\dfrac{n}{\mbox{Area}*P(\mbox{observing object}|\mbox{distance }r)}$

@@

Figure~\ref{fig:82dist} illustrates the distance data for the Collared Kingfisher. This data represents the aggregated data from two islands, Tinian and Rota, and all four observers. Most of the observations were made by sound, with a smaller percent visual, and a smaller percent heard-then-seen. We can observe several things in this plot:

@@

Secondly, the further away from the station, the distance observations were heaped into convenient distances. With smaller bin widths, this is also evident in the 100--200 m range as well. See Figure~\ref{fig:by5}.

Lastly, note that the detection distances are high between 100--200 m, and recall that the stations were placed 150 m apart. Since the stations on a transect were generally visited in sequence, this means it is possible a bird could have been observed from more than one station. How does this affect the independence of our observations? Will it make a difference in our final population density estimates if a bird is observed from more than one station?

\subsection{Independence}
Both \textcite{ramsey1979,buckland1987} discuss the assumption that the VCPs are placed randomly through the study area. This would seem to implicitly state that their observational ranges might by chance overlap, meaning a single object of interest could be observed from more than one station.  \textcite{thompson2012} likewise indicates random placement of line transects which would seem to allow for a potential overlap of observation zones.

@@

\textcite{buckland2001} discuss that each line or point must be ``randomly and independently located,'' with ``an equal probability of selection'' for all portions of the study area.

\textcite{buckland2001} states ``Transects are normally spaced at a sufficient distance to avoid detecting an object from two neighboring transacts, although this is not usually critical unless sampling a line changes the animal distribution at neighboring, as yet unsampled lines.''

@@

In discussing line transects, \textcite{barry2001} treat it as a two-stage sample, with non-unique clusters (each study object can be in more than one cluster) having an equal probability of selection. This would seem to (but does not explicitly) endorse the possibility of overlapping VCPs.

One of the assumptions of distance sampling is that the objects being studied are distributed homogeneously with some uniform density, $D$, throughout the space. \parencite{ramsey1981} (If the data are clustered we have other problems, not addressed here.) If the uniform distribution is the case, then it would make sense that it should not matter if the observations overlap, as long as the act of observing at one station does not affect the observations at the neighboring station. This echoes Buckland’s statement a few paragraphs previous.

\textcite{thompson2012} mentions that the systematic selection of transects (random position of first, with additional transects a fixed distance apart) does not affect the approximate unbiasedness of the density estimator, but will have an effect on the approximate unbiasedness of the variance estimator.

It is clear the literature is divided on whether or not our observation areas should overlap. VCP layouts similar to \textcite{micronesian} ``random-systematic'' layout are common in the literature. We as scientists need to be aware of any bias this may introduce into our estimates.

\subsection{Bias \& Consistency}
\textcite{buckland2001,ramsey1979} both claim that if our detectability estimate $g(z)=$1 out to $r$, then $\hat{D}$ is unbiased.

@@

	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/layout_rand-sys-4.pdf}
		\caption{A Transect Layout}
		\label{fig:transect1}
	\end{subfigure}
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\textwidth]{../images/layout_rand-sys-6.pdf}
		\caption{Another Transect Layout}
		\label{fig:transect2}
	\end{subfigure}
	\caption{VCP Layout options. On graph, 1 unit = 1 km. Circles represent 200 m mark.}
	\label{fig:layouts}
\end{figure}
Figures~\ref{fig:structured} and \ref{fig:random} illustrate the first two arrangements. Density in these images is based on the 20 birds per km$^2$ of the Palie region on Rota. \parencite{micronesian}

As seen in Figure~\ref{fig:82dist} or Figure~\ref{fig:by5}, there is very little detection happening beyond 500 m. There are a total of 15 observations beyond 500 m, which is 0.9\% of the data. For ease of simulation, 500 m was designated as our truncation distance, $w$, the distance beyond which the probability of detection is effectively 0.

@@

For the simulations, 0.1 unit = 1 km. All figures except Figure~\ref{fig:82dist} and \ref{fig:by5} follow this convention.

@@

Using the \texttt{hist()} function in R, the 1982 observation distances were broken into counts by 5 m groups. Using the bin midpoints, the density estimate was multiplied by 5 to get the density value for that bin. A LOESS regression was run with the midpoint as $x$ and $y=density*5$, using the \texttt{loess} package in R with a span of 0.2. This is the solid line in Figure~\ref{fig:by5}.

This gives us an object that can be used with the \texttt{predict} function to get density estimates for specific distances. The values generated by feeding our midpoints into our prediction function are represented by the dashed line in Figure~\ref{fig:by5}.

This empirical detection function needs to be scaled as we did with the half normal detection function. A sequence of values was generated covering the range from 0--500 m. These were plugged into the \texttt{predict()} function using our LOESS object. The highest density value was then scaled to be 1:

@@

\subsubsection{Movement}
In Figure~\ref{fig:82dist}, the histogram of empirical observation distances, there appears to be evidence of movement away from the observer. To see if the movement would have an effect on the independence of our VCPs, depending on placement, a second simulation was run.

Three movement classes were defined:
\begin{enumerate}
\item No Movement: Objects stayed at original point.
\item Temporary Movement: Objects within 100 m of the observer had a random chance to move directly away from the observer. Movement was larger if the object started closer to observer. Objects were reset to original location before moving to next VCP
\item Permanent Movement: As with temporary movement, only object location was not reset before moving on to next VCP. (Movement could compound if object was within the maximum observation radius for more than one VCP.)
\end{enumerate}

@@

\section{Results}
Two simulations were run, one looking at the three layout options crossed with the two detection probability function options and the other examining the effect of the three layout options against the three movement options, holding the detection probability option standard.
\subsection{Half-normal vs. Empirical Detection Probability}
\begin{figure}

@@

\end{figure}
Figure \ref{fig:sim1} shows the results of 1000 simulations, each analyzed with Empirical and Half-normal detection function, $g(x)$, and using each of the three layouts, Structured, Random, and Transect. All $\hat{D}$ estimations were done using the kernel method with a normal kernel \parencite{quang1993}.

@@

As we would expect, the density estimates using $g(x)$ based on the Empirical distance observations are biased low. We also see bias in the estimates using the half-normal function for $g(x)$, but it is much less severe than those based on the Empirical data. \textcite{quang1993} does provide a bias-correction method, but it was not used for this simulation.

@@

Table~\ref{table:sim1} supports what we observe in the histogram: the centers vary between our two $g(r)$ options, but are approximately the same within the each option, regardless of which VCP layout we have chosen. Table~\ref{table:sim1capture} shows the percentage of simulations that captured the true object density of 20 per km$^2$ within a 95\% confidence interval. \parencite{quang1993} Again, it is fairly consistent within detection probability type, regardless of layout choice.

There may be evidence of a drop in accuracy with the Transect layout. Both the mean and median is lower for the transect layout (Table~\ref{table:sim1} and \ref{table:sim1capture}), but it is difficult to determine if this is a valid observation, or an artifact of this particular simulation. The drop, if it exists, is much less severe with the half-normal detection function.

With the empirical detection function, the structured layout percentage swings much higher than the result from our random layout, unlike for half-normal where they are just over 1\% apart. Repeated simulations, could help narrow down if this is a true effect, or simply random noise.
\begin{table}[h]

@@

\end{figure}

For testing combinations of movement and VCP layout, only the half-normal detection function was utilized. Figure~\ref{fig:sim2} compares the histograms of the nine combinations. On the whole, the estimates are biased low, as was observed in the previous simulation, without movement. Within a column, the results are fairly similar in shape and spread. The most obvious differences lie between columns, indicating that the type of movement causes more bias in our estimate than the potential overlap of our observation areas.

The combination of compounded movement and the transect layout seems to have an effect on the density estimate. The other estimates for either movement type, are all in the same ballpark, around 11.8 with an approximate 59\% rate for capturing the true population value. For the Transect Layout and Compounded movement, this increases by 0.50 to a mean population density estimate of 12.42, and a capture percent of 65\%. It is unclear if there is something about this combination that is reducing bias (possible, but feels unlikely), or if it is instead biasing the estimate upwards, which appears like bias reduction because the estimate is already low.

\begin{table}[h]

@@

@@

\section{The Problem with Circles}
To test the effectiveness of the movement algorithm, the detection distances from the data where the objects moved was compared to the detection distances from the same data, assuming the objects stayed still. The movement data, (left panel, Figure~\ref{fig:moveTest}) is squashed more towards 0.1 Km than the still data (right panel). But there is a problem: The still data does not show the nice shoulder at 0 that the literature instructs us to expect.

@@

The take-away point is that for a VCP survey, even if everything is perfect, the empirical detection distances will not show the nice shouldered layout of the ideal detection probability curve. \textcite{ramsey1981} speak about this, when comparing the ``CUM-D'' curve, or the plot of the cumulative number of observations at distance $r$ for both line and point transects. However, in the other references examined, almost all the illustrations on how the collected data should look, shows data that echoes the detection probability curve: the half-bell-shape with a nice shoulder. Since this shape is impossible for point transects, explicit examples should be shown in the literature so researchers do not assume there is something wrong with their data.

@@

There is a possible effect of the combination of compounded movement and the transect layout. It is unclear from these results what the particular cause of the increased estimate and accuracy is. Future work might involve modification of the simulation functions to return more information, such as the count of observed objects and the observation distances to assist with understanding why this combination is producing an estimate that appears out of line with it's neighbors.

@@

\printbibliography
%\bibliographystyle{plain}
%\bibliography{VCPbibliography}

\subsection*{Code}
Simulation code is available at \url{https://github.com/mhedwards/wombat-VCP}
\end{document}
